---
title: "Global Economy Redesign"
subtitle: "CSC3107 Information Visualization · Team Poster Project"
author: "Team Lavender"
date: today
format:
  html:
    toc: true
    code-fold: show
    df-print: paged
execute:
  warning: false
  message: false
---

## Overview

This visualization compares the world's top 15 economies by:

1. Total GDP (market size) — horizontal bar chart
2. GDP per capita (individual affluence) — horizontal bar chart
3. Regional GDP share trends (2000–2026) — line chart

**Data sources:** IMF World Economic Outlook (April 2026) and World Bank.
**Output:** Combined multi-panel figure saved as `gdp_redesign.png`.

# Setup

```{r}
#| label: setup
library(tidyverse)
library(scales)
library(patchwork)
theme_set(theme_minimal(base_size = 12))
```

::: {.callout-note appearance="simple" icon=false}
### **Design Brief**
**Audience:** Economic and trade policy analysts.

**Analytical Task:** Compare the total economic scale of the top 15 global economies, track regional trajectories over the past decade, and contrast aggregate market size with individual affluence.

**Design Rationale:**

- **Symmetric Horizontal Bars:** Aligned side-by-side to allow direct, eye-movement-only comparison of total GDP (market size) against GDP per capita (prosperity) for the exact same country rows.
- **Direct-Labeled Line Chart:** Illustrates the global regional shifts from 2016 to 2026, highlighting the dominant scale of Asia's aggregate economy without legend-lookup clutter.
- **Unified Region Colors:** A cohesive color-blind-friendly palette links regions across all bar, line, and KPI card panels.
- **Structured KPI Banner:** Synthesizes high-level macro statistics at the bottom for quick executive scanning.
:::

# Data Preparation

```{r}
#| label: data-prep

# Data Loading
gdp_2026_pc <- read_csv("data/gdp_2026_with_percapita.csv", show_col_types = FALSE)
gdp_long <- read_csv("data/gdp_clean_long.csv", show_col_types = FALSE)

# Color Palette (ColorBrewer Set1)
region_pal <- brewer_pal(palette = "Set1")
region_colors <- setNames(region_pal(5), c("Asia", "Americas", "Europe", "Oceania", "Africa"))

# Top 15 economies by total GDP, ordered for plotting
top_15 <- gdp_2026_pc |>
  slice_max(gdp_usd_tn, n = 15) |>
  mutate(country = fct_reorder(country, gdp_usd_tn)) |>
  mutate(region = factor(region, levels = names(region_colors)))

# Regional GDP share over time (2000–2026)
regional_trend <- gdp_long |>
  filter(!is.na(region), year >= 2000, year <= 2026) |>
  group_by(year, region) |>
  summarize(gdp_usd_tn = sum(gdp_usd_tn), .groups = "drop_last") |>
  mutate(share = gdp_usd_tn / sum(gdp_usd_tn))

# Order regions by 2026 GDP (largest first)
region_order <- regional_trend |>
  filter(year == 2026) |>
  arrange(desc(gdp_usd_tn)) |>
  pull(region)

regional_trend <- regional_trend |>
  mutate(region = factor(region, levels = region_order))

```

## Panel A: Market Size (Total GDP)

The United States and China dominate the global landscape, together accounting for nearly half of the top 15's total output. While the US maintains the top position at $32.4T, China follows closely at $20.9T, with a significant drop-off to the next tier of economies like Germany and Japan.

```{r}
#| label: plot-a
plot_a <- ggplot(top_15, aes(x = gdp_usd_tn, y = country, fill = region)) +
  geom_col(width = 0.75, show.legend = FALSE) +
  geom_text(
    aes(label = sprintf("$%.1fT", gdp_usd_tn)),
    hjust = -0.15,
    color = "gray20",
    fontface = "bold",
    size = 3.2
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.15)),
    labels = label_dollar(suffix = "T")
  ) +
  scale_fill_manual(values = region_colors) +
  labs(
    title = "Panel A: Market Size (Nominal GDP)",
    subtitle = "Top 15 economies in 2026 (USD Trillions)",
    x = "GDP in 2026",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(color = "gray40", size = 10),
    axis.text.y = element_text(face = "bold", color = "gray20", size = 10)
  )

plot_a
```

## Panel B: Individual Affluence (GDP per capita)

Contrasting market size with individual affluence reveals a sharp divide. While the US and European nations maintain high GDP per capita (prosperity), the emerging giants China and India show significantly lower per-capita figures. This highlights that massive market scale does not yet equate to the same level of individual wealth found in established advanced economies.

```{r}
#| label: plot-b
plot_b_affluence <- ggplot(top_15, aes(x = gdp_per_capita, y = country, fill = region)) +
  geom_col(width = 0.75, show.legend = FALSE) +
  geom_text(
    aes(label = sprintf("$%.1fk", gdp_per_capita / 1000)),
    hjust = -0.15,
    color = "gray20",
    fontface = "bold",
    size = 3.2
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.2)),
    labels = label_dollar(scale = 0.001, suffix = "k")
  ) +
  scale_fill_manual(values = region_colors) +
  labs(
    title = "Panel B: Individual Affluence",
    subtitle = "GDP per capita in 2026 (USD)",
    x = "GDP per capita",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(color = "gray40", size = 10),
    axis.text.y = element_blank()
  )

plot_b_affluence
```

## Panel C: Regional GDP Share Trends

The "Shift to Asia" is most evident when examining regional shares over time. Since 2000, Asia's share of global GDP has risen from under 20% to over 30%, surpassing both the Americas and Europe. This structural transformation underscores the migration of global economic gravity toward the East.

```{r}
#| label: plot-c
# Extract 2026 data for direct labels
label_data <- regional_trend |>
  filter(year == 2026)

plot_c_trends <- ggplot(regional_trend, aes(x = year, y = share, color = region, group = region)) +
  geom_line(linewidth = 1.2, show.legend = FALSE) +
  geom_point(data = label_data, size = 2.0, show.legend = FALSE) +
  scale_x_continuous(
    limits = c(2000, 2027.5),
    breaks = seq(2000, 2025, by = 5),
    labels = c("2000", "2005", "2010", "2015", "2020", "2026"),
    expand = expansion(mult = c(0.02, 0))
  ) +
  scale_y_continuous(
    labels = label_percent(),
    breaks = seq(0, 0.4, by = 0.1)
  ) +
  scale_color_manual(values = region_colors) +
  labs(
    title = "Panel C: Global GDP Share by Region (2000-2026)",
    subtitle = "Regional share of total GDP; Asia's share has grown significantly",
    x = "Year",
    y = "Share of World GDP"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(color = "gray40", size = 10)
  )

plot_c_trends
```

## Final Dashboard Assembly

The final figure combines all insights into a single unified dashboard, facilitating multi-dimensional analysis of global economic power.

```{r}
#| label: dashboard-combine
#| fig-width: 14
#| fig-height: 6.5

# Legend plot (one row per region for the legend keys)
legend_data <- tibble(region = names(region_colors))
region_legend <- ggplot(legend_data, aes(x = 1, y = 1, fill = region)) +
  geom_tile(alpha = 0) +
  scale_fill_manual(values = region_colors, drop = FALSE) +
  guides(fill = guide_legend(nrow = 1, title = "Region",
    title.position = "top", override.aes = list(alpha = 1))) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.key.spacing.x = unit(0.8, "cm"),
    legend.title = element_text(hjust = 0.5),
    legend.margin = margin(t = 3, b = 3),
    plot.margin = margin(0, 0, 0, 0)
  )

combined_plot <- plot_a + plot_b_affluence + plot_c_trends + region_legend +
  plot_layout(
    design = "
      AAABBBCCC
      AAABBBCCC
      ##DDDDD##
    ",
    heights = c(1, 1, 0.14)
  ) +
  plot_annotation(
    title = "Redesigning the Global Economy: Market Size, Trends, and Individual Affluence",
    subtitle = "A comparison of the world's top 15 economies (2000-2026)",
    caption = "Data Source: IMF World Economic Outlook (April 2026) & World Bank Population Data.",
    theme = theme(
      plot.title = element_text(size = 18, face = "bold", hjust = 0, margin = margin(b = 5)),
      plot.subtitle = element_text(size = 12, hjust = 0, color = "gray30", margin = margin(b = 20)),
      plot.caption = element_text(size = 9, color = "gray50", margin = margin(t = 15))
    )
  )

ggsave("gdp_redesign.png", combined_plot, width = 14, height = 6.5, dpi = 300)

combined_plot
```

