---
title: "The Global Economy Beyond a Snapshot: Concentration, the Shift to Asia, and Prosperity"
subtitle: "CSC3107 Information Visualization · Team Poster Project"
author: "Team [your team ID]"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: show
    df-print: paged
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

```{r setup}
#| label: setup
# Install once (all on CRAN):
# install.packages(c("tidyverse","readxl","countrycode","WDI","scales","patchwork"))
library(tidyverse)   # dplyr, tidyr, readr, ggplot2, stringr, forcats
library(readxl)      # read the IMF DataMapper .xls directly (no Excel step)
library(countrycode) # ISO-3 codes + region
library(WDI)         # World Bank population (downloaded in-code) for GDP per capita
library(scales)
library(patchwork)   # compose multiple ggplots into panelled figures (in code)
theme_set(theme_minimal(base_size = 12))

# colour-blind-safe (Okabe-Ito): one distinct hue per region, used consistently
region_pal <- c("Africa"="#E69F00","Asia"="#0072B2","Europe"="#009E73",
                "Americas"="#D55E00","Oceania"="#CC79A7","Rest of World"="#999999")
```

# 1. The problem we are solving

The original graphic (Visual Capitalist, *The \$126T Global Economy in One Chart*) is a
single-year **2026 treemap**. Its accompanying text claims economic weight is **shifting
toward Asia** — but a one-year snapshot is structurally incapable of showing *change*. The
chart asserts a trend it cannot draw.

Our redesign, for **economic and policy analysts**, answers three questions the original
cannot:

1. **Concentration** — is global output still dominated by a handful of economies?
2. **The shift** — is economic weight moving between regions, specifically toward Asia?
3. **Prosperity** — among the largest economies, which are actually affluent *per person*,
   and who is catching up?

The data confirm the story is real: Asia's share of world GDP rose from **22.9% (1980) to
35.3% (2026)**, overtaking Europe (which fell from 35.5% to 25.7%) and the Americas.

# 2. Data source

Nominal GDP: **IMF World Economic Outlook, April 2026**, indicator **`NGDPD`** (*GDP,
current prices, USD billions*), exported from the IMF DataMapper as
`data/imf-dm-export-20260624.xls`. Population: **World Bank `SP.POP.TOTL`**, downloaded
in-code via `WDI` and merged to derive GDP per capita. **All processing happens in this
document** — the raw `.xls` is read directly with `readxl`; nothing is edited by hand.

# 3. Import & clean the IMF GDP data

The export lists individual economies first (Afghanistan → Zimbabwe), then regional and
analytical aggregates (`Africa`, `Advanced economies`, `G-20`, …, `World`) and a footer.
Summing everything double-counts, so we keep only individual economies: cut at the first
aggregate, then validate names against ISO-3 (dropping anything that does not resolve).

```{r clean}
#| label: clean
raw <- read_excel("data/imf-dm-export-20260624.xls") |> rename(entity = 1)

first_agg <- which(raw$entity == "Africa")[1]         # first regional aggregate
econ <- raw |>
  slice(1:(first_agg - 1)) |>
  filter(!is.na(entity)) |>
  select(-any_of("Estimates start after"))

year_cols <- names(econ)[str_detect(names(econ), "^[0-9]{4}$")]
custom <- c("Türkiye, Republic of"="TUR","Kosovo"="XKX",
            "Micronesia, Fed. States of"="FSM","Lao P.D.R."="LAO")

gdp_long <- econ |>
  pivot_longer(all_of(year_cols), names_to = "year", values_to = "gdp_usd_bn") |>
  mutate(
    year       = as.integer(year),
    gdp_usd_bn = parse_number(na_if(as.character(gdp_usd_bn), "no data")),
    iso3       = countrycode(entity, "country.name", "iso3c", custom_match = custom, warn = FALSE),
    country    = countrycode(iso3, "iso3c", "country.name", warn = FALSE),
    region     = countrycode(iso3, "iso3c", "continent",    warn = FALSE)  # Africa/Americas/Asia/Europe/Oceania
  ) |>
  filter(!is.na(gdp_usd_bn), !is.na(iso3)) |>
  mutate(gdp_usd_tn = gdp_usd_bn / 1000) |>
  select(iso3, country, region, year, gdp_usd_bn, gdp_usd_tn)

dir.create("data", showWarnings = FALSE)
write_csv(gdp_long, "data/gdp_clean_long.csv")

gdp_2026 <- gdp_long |> filter(year == 2026) |> select(-year) |> arrange(desc(gdp_usd_tn))
write_csv(gdp_2026, "data/gdp_clean_2026.csv")

sprintf("Economies: %d | 2026 total: $%.1fT (IMF World = $126.3T)",
        n_distinct(gdp_long$iso3), sum(gdp_2026$gdp_usd_tn))
```

# 4. Merge population -> GDP per capita

World Bank population is joined on `iso3` + `year`, giving GDP per capita for every
country-year (the merge your project requires). World Bank population is historical, so the
per-capita series runs to the latest available year (~2024); GDP-only analyses use the full
1980–2031 range.

```{r population}
#| label: population
pop_long <- WDI(indicator = c(population = "SP.POP.TOTL"),
                start = 1980, end = 2026, extra = FALSE) |>
  as_tibble() |>
  filter(!is.na(population)) |>
  mutate(iso3 = countrycode(iso2c, "iso2c", "iso3c", warn = FALSE)) |>
  filter(!is.na(iso3)) |>
  select(iso3, year, population)

# per-capita panel data (all years where both GDP and population exist)
pc_long <- gdp_long |>
  inner_join(pop_long, by = c("iso3", "year")) |>
  mutate(gdp_per_capita = gdp_usd_tn * 1e12 / population)

# 2026 per-capita snapshot
gdp_2026_pc <- gdp_2026 |>
  left_join(pop_long |> group_by(iso3) |> slice_max(year, n = 1, with_ties = FALSE) |>
              ungroup() |> select(iso3, population), by = "iso3") |>
  mutate(gdp_per_capita = gdp_usd_tn * 1e12 / population)
write_csv(gdp_2026_pc, "data/gdp_2026_with_percapita.csv")
```

# 5. The four analyses

The 15 largest economies (by 2026 GDP) are the focus set your instructor suggested.

```{r top15}
#| label: top15
top15 <- gdp_2026 |> slice_max(gdp_usd_tn, n = 15)
top15_iso <- top15$iso3
```

## 5.1 Panel A — The shift: regional share of world GDP over time

A **line chart** (not bars) is the right tool for change over time: it shows Asia's rise
and the crossover with Europe/Americas directly. This is the story the original snapshot
could not tell.

```{r fig-shift}
#| label: fig-shift
#| fig-cap: "Regional share of world GDP, 1980-2031. Asia overtakes Europe and the Americas."
world_by_year <- gdp_long |> group_by(year) |> summarise(world = sum(gdp_usd_bn), .groups = "drop")
regional_share <- gdp_long |>
  group_by(year, region) |>
  summarise(region_gdp = sum(gdp_usd_bn), .groups = "drop") |>
  left_join(world_by_year, by = "year") |>
  mutate(share = region_gdp / world * 100)

end_labels <- regional_share |> filter(year == max(year))

p_shift <- ggplot(regional_share, aes(year, share, colour = region)) +
  geom_line(linewidth = 1.1) +
  geom_text(data = end_labels, aes(label = region), hjust = 0, nudge_x = 0.4,
            size = 3.4, fontface = "bold", show.legend = FALSE) +
  scale_colour_manual(values = region_pal, guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.14))) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(title = "Economic weight is shifting to Asia",
       subtitle = "Asia's share of world GDP rose from 23% (1980) to 35% (2026)",
       x = NULL, y = "Share of world GDP")

p_shift
```

## 5.2 Panel B — Concentration today: top 15 + Rest of World

```{r fig-concentration}
#| label: fig-concentration
#| fig-cap: "Nominal GDP, 2026: the 15 largest economies plus an aggregated Rest of World."
row_val <- sum(gdp_2026$gdp_usd_tn) - sum(top15$gdp_usd_tn)
conc <- top15 |>
  select(country, region, gdp_usd_tn) |>
  bind_rows(tibble(country = "Rest of World", region = "Rest of World", gdp_usd_tn = row_val)) |>
  mutate(country = fct_reorder(country, gdp_usd_tn),
         country = fct_relevel(country, "Rest of World"))

top4_pct <- round(100 * sum(sort(gdp_2026$gdp_usd_tn, decreasing = TRUE)[1:4]) / sum(gdp_2026$gdp_usd_tn))

p_conc <- ggplot(conc, aes(gdp_usd_tn, country, fill = region)) +
  geom_col(width = 0.75) +
  geom_text(aes(label = number(gdp_usd_tn, accuracy = 0.1)), hjust = -0.15, size = 3.2) +
  scale_fill_manual(values = region_pal, name = "Region") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12)), labels = label_number(suffix = "T")) +
  labs(title = "A few economies dominate world output",
       subtitle = glue::glue("The top 4 produce about {top4_pct}% of world GDP"),
       x = "Nominal GDP, 2026 (USD trillions)", y = NULL) +
  theme(legend.position = "bottom")

p_conc
```

## 5.3 Panel C — Prosperity now: GDP per capita of the largest economies

Ordering the same 15 economies by **GDP per capita** shows that size is not prosperity —
China and India are huge in total but low per person.

```{r fig-percap-now}
#| label: fig-percap-now
#| fig-cap: "GDP per capita, 2026, for the 15 largest economies."
pc_now <- gdp_2026_pc |>
  filter(iso3 %in% top15_iso) |>
  mutate(country = fct_reorder(country, gdp_per_capita))

p_percap <- ggplot(pc_now, aes(gdp_per_capita, country, fill = region)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = dollar(gdp_per_capita, accuracy = 1, scale = 1e-3, suffix = "k")),
            hjust = -0.12, size = 3.2) +
  scale_fill_manual(values = region_pal, name = "Region") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.16)), labels = label_dollar(scale = 1e-3, suffix = "k")) +
  labs(title = "Size is not prosperity",
       subtitle = "The largest economies, ranked by output per person (2026)",
       x = "GDP per capita (USD)", y = NULL) +
  theme(legend.position = "bottom")

p_percap
```

## 5.4 Panel D — Catch-up: GDP-per-capita trends as small multiples

For 15 trajectories, **faceting** (`facet_wrap`) generates one panel per economy *in code*
— no copy-paste. Free y-scales let each economy's growth shape show; this reveals which are
catching up (steep Asian risers) versus mature and flat.

```{r fig-percap-trend}
#| label: fig-percap-trend
#| fig-height: 7
#| fig-cap: "GDP per capita over time, one small-multiple panel per economy (top 15)."
pc_trend <- pc_long |>
  filter(iso3 %in% top15_iso) |>
  mutate(country = fct_reorder(country, gdp_per_capita, .fun = max, .desc = TRUE))

p_trend <- ggplot(pc_trend, aes(year, gdp_per_capita, colour = region)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = region_pal, guide = "none") +
  scale_y_continuous(labels = label_dollar(scale = 1e-3, suffix = "k")) +
  facet_wrap(~ country, scales = "free_y", ncol = 5) +
  labs(title = "Who is catching up? GDP per capita over time",
       x = NULL, y = "GDP per capita (USD)")

p_trend
```

## 5.5 Summary statistics

```{r summary-stats}
#| label: summary-stats
# (a) regional share, 1980 vs 2026, and the change
share_change <- regional_share |>
  filter(year %in% c(1980, 2026)) |>
  select(region, year, share) |>
  pivot_wider(names_from = year, values_from = share, names_prefix = "yr_") |>
  mutate(change_pts = yr_2026 - yr_1980) |>
  arrange(desc(yr_2026))
share_change

# (b) per-capita compound annual growth rate (CAGR) for the top 15
pc_cagr <- pc_trend |>
  group_by(country, region) |>
  summarise(y0 = min(year), y1 = max(year),
            pc0 = gdp_per_capita[which.min(year)],
            pc1 = gdp_per_capita[which.max(year)], .groups = "drop") |>
  mutate(cagr = scales::percent((pc1 / pc0)^(1 / (y1 - y0)) - 1, accuracy = 0.1)) |>
  arrange(desc((pc1 / pc0)^(1 / (y1 - y0)) - 1)) |>
  select(country, region, from = y0, to = y1, cagr)
pc_cagr
```

## 5.6 (Optional) Compose panels into one figure

For the poster you will lay panels out yourself, but `patchwork` can also assemble them in
code — the sanctioned alternative to copy-pasting separate images.

```{r compose}
#| label: fig-compose
#| eval: false
(p_shift | p_conc) / (p_percap | p_trend) +
  plot_annotation(title = "The global economy beyond a snapshot")
```

# 6. How this helps the audience (why each panel earns its place)

For **economic and policy analysts** deciding where to focus trade, investment, and
diplomatic resources:

- **Panel A (the shift)** tells them *where future demand is concentrating*. A sustained
  rise in Asia's share is a signal to prioritise Asian markets and partnerships now, ahead
  of the trend — the single most actionable insight, and the one the original chart hides.
- **Panel B (concentration)** quantifies *exposure and dependency*: because a few economies
  drive global demand, a shock in any of them propagates worldwide. This informs risk
  assessment and the case for diversification.
- **Panel C (prosperity)** separates *market size from purchasing power*. A large economy
  (China, India) is a high-*volume* market; a wealthy one (US, Germany) is a high-*value*
  one. This tells an analyst whether to target a market for scale or for premium goods.
- **Panel D (catch-up)** flags *emerging affluent consumers*: economies with steep
  per-capita growth are tomorrow's high-value markets, supporting early market-entry
  decisions.

Together they answer a question the snapshot cannot: not just *who is largest today*, but
*who is rising, who is wealthy, and who is catching up*.

# 7. Choice of time window (and why)

The IMF WEO series runs **1980–2031**. We show the **full range** for the regional-share and
per-capita-trend charts because the *long arc is the point*: a structural shift like Asia's
rise (+12 points over four decades) is only credible when the whole trajectory is visible,
and 1980 is the start of the IMF's internally consistent dataset. Analysts commonly use
either the full available series (to show structural change) or a ~20–30-year window (to
reduce clutter); for the specifically *modern* Asia story, the acceleration is post-2000
(around China's 2001 WTO accession), so a 2000-onward view is a defensible sharper cut. The
start year is a single parameter — set `start_year <- 2000` and filter — so you can switch
views without changing anything else.

```{r time-window}
#| label: time-window
#| eval: false
start_year <- 2000
regional_share_recent <- regional_share |> filter(year >= start_year)
# re-plot p_shift with regional_share_recent to show the modern-era view
```

# 8. Cleaning & transformation log (for the report / poster "Data" panel)

1. **Read raw `.xls` in R** (`readxl`) — no spreadsheet conversion.
2. **Removed aggregates** — cut at the first regional row (`Africa`); validated names with
   `countrycode`, dropping anything not resolving to ISO-3, plus empty rows and the footer.
3. **Reshaped** wide year columns to tidy long; parsed `"no data"` to `NA`; dropped missing.
4. **Standardised** identifiers (ISO-3, small `custom_match`) and units (billions → trillions).
5. **Merged World Bank population** (`SP.POP.TOTL`) on `iso3` + `year`; derived GDP per capita.
6. **Derived** regional shares of world GDP, top-15 concentration, per-capita levels and
   trends, and summary statistics (share change 1980→2026; per-capita CAGR).

**Result:** 190 economies for 2026 (\$125.7T vs IMF World \$126.3T); a verified rise in
Asia's share from 22.9% to 35.3% (1980→2026).
